### 2025-09-14 YEAR2/AER210_LectureNotes/main.tex
push

\subsection{Applications of Double Integrals}
\subsection{Surface Area}

>>>>

\begin{example}
    \textit{Find the area of the region $R$ enclosed by one petal of the rose given by $r = \cos{3\theta}$.}

    We know that the petal has upper bound at $\theta = \pm \frac{\pi}{6}$. We can describe the region $R$ in polar coordinates as $0 \le r \le \cos{3\theta}$ and $-\frac{\pi}{6} \le \theta \le \frac{\pi}{6}$. Thus, we can set up the double integral as follows:
    \begin{align*}
        \text{Area}(R) &= \iint_R 1 \, dA \\
        &= \int_{-\frac{\pi}{6}}^{\frac{\pi}{6}} \int_0^{\cos{3\theta}} r \, dr \, d\theta \\
        &= \int_{-\frac{\pi}{6}}^{\frac{\pi}{6}} \left( \frac{r^2}{2} \Big|_{r=0}^{r=\cos{3\theta}} \right) d\theta \\
        &= \int_{-\frac{\pi}{6}}^{\frac{\pi}{6}} \frac{\cos^2{3\theta}}{2} \, d\theta = \frac{\pi}{12} \quad \left(\text{Using } \cos^2{x} = \frac{1 + \cos{2x}}{2}\right)
    \end{align*}
\end{example}

\begin{example}
    \textit{Find the volume trapped between the cone $z = \sqrt{x^2 + y^2}$ and the sphere $x^2 + y^2 + z^2 = 1$.}

    First, we find the intersection of the cone and the sphere:
    $$
        z = \sqrt{x^2 + y^2} \implies z^2 = x^2 + y^2
    $$
    Substituting into the sphere equation:
    \begin{align*}
        x^2 + y^2 + z^2 = 1 &\implies z^2 + z^2 = 1 \\ 
        &\implies 2z^2 = 1 \implies z = \frac{1}{\sqrt{2}} \quad x^2 + y^2 = \frac{1}{2}
    \end{align*}
    Thus, we have the region $R = \{(r, \theta) \mid 0 \le r \le \frac{1}{\sqrt{2}}, 0 \le \theta \le 2\pi\}$. We can set up the double integral as follows:
    \begin{align*}
        V &= \iint_R (\sqrt{1 - x^2 - y^2} - \sqrt{x^2 + y^2} ) \, dA \\ % the shpere - cone
        &= \int_0^{2\pi} \int_0^{\frac{1}{\sqrt{2}}} (\sqrt{1 - r^2} - r) \, r \, dr \, d\theta \\
        &= 2\pi \left[ \int_0^{\frac{1}{\sqrt{2}}} r\sqrt{1 - r^2} \, dr - \int_0^{\frac{1}{\sqrt{2}}} r^2 dr \right] \\
        &= 2\pi \left[ -\frac{1}{3} (1 - r^2)^{\frac{3}{2}} \Big|_0^{\frac{1}{\sqrt{2}}} - \frac{r^3}{3} \Big|_0^{\frac{1}{\sqrt{2}}} \right] = \frac{2\pi}{3} (1- \frac{1}{\sqrt{2}})
    \end{align*}
\end{example}

\subsection{Applications of Double Integrals}
\begin{shaded}
    \subsubsection*{Review: Moment of Inertia using Single Integral}

    Consider a thin rod of length $L$ with a linear mass density $\rho(x)$, where $x$ is the distance from one end of the rod. We can use the following formula to find the moment of inertia $I$ of the rod about an axis perpendicular to the rod and passing through one end:

    \begin{equation}
        I = \int_0^L x^2 \rho(x) \, dx
    \end{equation}
    where $x^2$ is the square of the distance from the axis of rotation, and $\rho(x) \, dx$ represents the mass element of the rod at position $x$. This comes from the definition of moment of inertia, which is the following sum with point masses:
    $$        
    I = \sum m_i r_i^2 \quad \text{s.t. } \quad \text{KE}= \frac{1}{2} I \omega^2
    $$
\end{shaded}

\begin{definition}[Mass of a Lamina]
    Consider a lamina occupying the region $R$ in the $xy$-plane with a surface mass density $\sigma(x,y)$, where $\sigma(x,y)$ is the mass per unit area at the point $(x,y)$. The mass $M$ of the lamina can be found using the following double integral:
    \begin{equation}
        M = \iint_R \sigma(x,y) \, dA
    \end{equation}
    where $dA$ represents an infinitesimal area element in the region $R$.

\end{definition}

\begin{definition}[Moment of a Lamina]
    The moment of the lamina about the $x$-axis ($M_x$) and $y$-axis ($M_y$) can be found using the following formulas:
    \begin{equation}
        M_x = \iint_R y \sigma(x,y) \, dA, \quad M_y = \iint_R x \sigma(x,y) \, dA
    \end{equation}
    where $y$ and $x$ are the distances from the respective axes of rotation, and $\sigma(x,y) \, dA$ represents the mass element of the lamina at the point $(x,y)$.
\end{definition}

\begin{definition}[Center of Mass of a Lamina]
    The center of mass $(\bar{x}, \bar{y})$ of the lamina can be found using the following formulas:
    \begin{equation}
        \bar{x} = \frac{1}{M} M_y = \frac{1}{M} \iint_R x \sigma(x,y) \, dA, \quad \bar{y} = \frac{1}{M} M_x = \frac{1}{M} \iint_R y \sigma(x,y) \, dA
    \end{equation}
    where $M$ is the total mass of the lamina as calculated in the previous example.
\end{definition}

\begin{example}
    \textit{Find the cnetre of mass of the following plate with density funciton $\sigma(x,y) = x + y$ over the reigion bounded by the axis and $y = \sqrt{x}$ and $x = 1$.}

    First, we find the mass of the lamina:
    \begin{align*}
        M &= \iint_R (x + y) \, dA \\
        &= \int_0^1 \int_0^{\sqrt{x}} (x + y) \, dy \, dx \\
        &= \int_0^1 \left[ xy + \frac{y^2}{2} \right]_{y=0}^{y=\sqrt{x}} \, dx \\
        &= \int_0^1 \left( x\sqrt{x} + \frac{x}{2} \right) dx = \left[ \frac{2}{5} x^{\frac{5}{2}} + \frac{1}{4} x^2 \right]_{0}^{1} = \frac{13}{20}
    \end{align*}
    Now, we find the moments about the $x$-axis and $y$-axis:
    \begin{align*}
        M_x &= \iint_R y(x + y) \, dA \\
        &= \int_0^1 \int_0^{\sqrt{x}} y(x + y) \, dy \, dx \\
        &= \int_0^1 \left[ \frac{xy^2}{2} + \frac{y^3}{3} \right]_{y=0}^{y=\sqrt{x}} \, dx \\
        &= \int_0^1 \left( \frac{x^2}{2} + \frac{x^{\frac{3}{2}}}{3} \right) dx = \left[ \frac{x^3}{6} + \frac{2}{15} x^{\frac{5}{2}} \right]_{0}^{1} = \frac{3}{10}
    \end{align*}
    and,
    \begin{align*}
        M_y &= \iint_R x(x + y) \, dA \\
        &= \int_0^1 \int_0^{\sqrt{x}} x(x + y) \, dy \, dx \\
        &= \int_0^1 \left[ x^2y + \frac{xy^2}{2} \right]_{y=0}^{y=\sqrt{x}} \, dx \\
        &= \int_0^1 \left( x^2\sqrt{x} + \frac{x^{\frac{3}{2}}}{2} \right) dx = \left[ \frac{2}{7} x^{\frac{7}{2}} + \frac{1}{6} x^3 \right]_{0}^{1} = \frac{19}{42}
    \end{align*}
    Finally, we can find the center of mass:
    $$        
    \bar{x} = \frac{M_y}{M} = \frac{190}{273}, \quad \bar{y} = \frac{M_x}{M} = \frac{6}{13}
    $$
\end{example}

\begin{definition}[Geometric Center of a Lamina]
    If the surface mass density $\sigma(x,y)$ is constant, then the center of mass is also known as the geometric center (or centroid) of the lamina. The formulas for the geometric center are:
    \begin{equation}
        x_c = \frac{1}{\text{Area}(R)} \iint_R x \, dA, \quad y_c = \frac{1}{\text{Area}(R)} \iint_R y \, dA
    \end{equation}
    where $\text{Area}(R)$ is the area of the region $R$.

\end{definition}

\begin{definition}[Moment of Inertia of a Lamina]
    The moment of inertia of the lamina about the $x$-axis ($I_x$) and $y$-axis ($I_y$) can be found using the following formulas:
    \begin{equation}
        I_x = \iint_R y^2 \sigma(x,y) \, dA, \quad I_y = \iint_R x^2 \sigma(x,y) \, dA
    \end{equation}
    where $y^2$ and $x^2$ are the squares of the distances from the respective axes of rotation, and $\sigma(x,y) \, dA$ represents the mass element of the lamina at the point $(x,y)$. Also for the moment of inertia about the origin ($I_o$):
    \begin{equation}
        I_o = \iint_R (x^2 + y^2) \sigma(x,y) \, dA
    \end{equation}
    In general, for an axis defined by a line $ax + by + c = 0$, the moment of inertia about that axis ($I_l$) can be found using the following formula:
    \begin{equation}
        I_l = \iint_R \left( \frac{ax + by + c}{\sqrt{a^2 + b^2}} \right)^2 \sigma(x,y) \, dA
    \end{equation}
\end{definition}

\begin{example}
    A rectangular plate of mass $m$, length $L$ and width $W$ is rotated about a vertical line on its left side with width $W$. Find the moment of inertia of the plate about this line in two cases:
    \begin{enumerate}
        \item The plate has uniform density $\sigma(x,y) = \frac{m}{LW}$.
        \item The density varies at a point proportional to the square of the distance from the right most side.
        \item It has uniform density, but rotated its center.
    \end{enumerate}

   \textbf{Solution. 1. Uniform Density} We can describe the region $R$ in Cartesian coordinates as $0 \le x \le L$ and $0 \le y \le W$. The surface mass density is $\sigma(x,y) = \frac{m}{LW}$. Thus, we can set up the double integral as it rotates around the $y$-axis:
    \begin{align*}
        I_y &= \iint_R x^2 \sigma(x,y) \, dA \\
        &= \int_0^W \int_0^L x^2 \cdot \frac{m}{LW} \, dx \, dy \\
        &= \int_0^W \left[ \frac{m}{LW} \cdot \frac{x^3}{3} \right]_{x=0}^{x=L} dy = \int_0^W \frac{mL^2}{3W} dy = \left[ \frac{mL^2}{3W} y \right]_{y=0}^{y=W} = \frac{mL^2}{3}
    \end{align*}
    \textbf{Solution. 2. Varying Density} We can describe the region $R$ in Cartesian coordinates as $0 \le x \le L$ and $0 \le y \le W$. The surface mass density is $\sigma(x,y) = k(L - x)^2$. To find the constant $k$, we have:
    \begin{align*}
        I_y &= \iint_R x^2 \sigma(x,y) \, dA  \\
        &= \int_0^W \int_0^L x^2 \cdot k(L - x)^2 \, dx \, dy \\
        &= \int_0^W \left[ k \int_0^L (x^2L^2 - 2Lx^3 + x^4) \, dx \right] dy \\
        &= \int_0^W \left[ k \left( \frac{L^5}{3} - \frac{L^5}{2} + \frac{L^5}{5} \right) \right] dy = \int_0^W \left[ k \cdot \frac{L^5}{30} \right] dy = k \cdot \frac{L^5}{30} W
    \end{align*}

    \textbf{Solution. 3. Rotated Center} We can describe the region $R$ in Cartesian coordinates as $-\frac{L}{2} \le x \le \frac{L}{2}$ and $-W \le y \le W$. The surface mass density is $\sigma(x,y) = \frac{m}{LW}$. Thus, we can set up the double integral as it rotates around the origin:
    \begin{equation*}
    \begin{split}
        I_o &= \iint_R (x^2 + y^2) \sigma(x,y) \, dA \\
        &= \int_{-W}^W \int_{-L/2}^{L/2} (x^2 + y^2) \cdot \frac{m}{LW} \, dx \, dy \\
        &= \int_{-W}^W \left[ \frac{m}{LW} \left( \frac{x^3}{3} + y^2 x \right)_{x=-L/2}^{x=L/2} \right] dy \\
        &= \int_{-W}^W \left[ \frac{mL^2}{12W} + \frac{my^2}{W} \right] dy \\
        &= \left[ \frac{mL^2}{12W} y + \frac{my^3}{3W} \right]_{y=-W}^{y=W} = \frac{mL^2}{6} + \frac{2mW^2}{3}
    \end{split}
    \end{equation*}
\end{example}

\subsection{Surface Area}
\begin{theorem}[Surface Area]
    Given $z=f(x,y)$, where $f$ is a differentiable function over the region $R$ in the $xy$-plane, the surface area $S$ of the surface above the region $R$ is given by:
    \begin{equation}
        S = \iint_R \sqrt{1 + \left( \frac{\partial f}{\partial x} \right)^2 + \left( \frac{\partial f}{\partial y} \right)^2} \, dA
    \end{equation}
\end{theorem}
\begin{proof}
    Consider a small rectangle $\Delta A_i$ in the $xy$-plane with dimensions $\Delta x_i$ and $\Delta y_i$. When we project this rectangle onto the surface $z = f(x,y)$, it becomes a parallelogram $T_i$ tangent to the surface. Using the partial derivatives of the surface, we can deduce the two vectors that define the parallelogram:
    $$
        \vec{u} = \left( \Delta x_i, 0, f_x(x_i^*, y_i^*) \Delta x_i \right), \quad \vec{v} = \left( 0, \Delta y_i, f_y(x_i^*, y_i^*) \Delta y_i \right)
    $$
    The area of this parallelogram is given by the magnitude of the cross product of these two vectors:
    \begin{align*}
        \Delta T_i = ||\vec{u} \times \vec{v}|| &= \left|\begin{vmatrix}
        \hat{i} & \hat{j} & \hat{k} \\
        \Delta x_i & 0 & f_x(x_i^*, y_i^*) \Delta x_i \\
        0 & \Delta y_i & f_y(x_i^*, y_i^*) \Delta y_i
        \end{vmatrix}\right| \\
        &= \sqrt{( -f_x(x_i^*, y_i^*) \Delta x_i \Delta y_i )^2 + ( -f_y(x_i^*, y_i^*) \Delta x_i \Delta y_i )^2 + (\Delta x_i \Delta y_i)^2} \\
        &= \sqrt{1 + (f_x(x_i^*, y_i^*))^2 + (f_y(x_i^*, y_i^*))^2} \, \Delta x_i \Delta y_i
    \end{align*}
    Therefore, the surface area can be approximated as:
    $$
        S \approx S_n = \sum_{i=1}^n \Delta T_i = \sum_{i=1}^n \sqrt{1 + (f_x(x_i^*, y_i^*))^2 + (f_y(x_i^*, y_i^*))^2} \, \Delta A_i
    $$
    Taking the limit as the maximum diameter of the subrectangles approaches zero, we obtain the exact surface area:
    $$
        S = \lim_{|P| \to 0} S_n =\iint_R \sqrt{1 + \left( \frac{\partial f}{\partial x} \right)^2 + \left( \frac{\partial f}{\partial y} \right)^2} \, dA
    $$
    which completes our derivation.
\end{proof}

\begin{example}
    \textit{Find the surface of the sphere $x^2 + y^2 + z^2 = a^2$. }

    We first consider the first octant, where $x, y, z \ge 0$. Then the total volumn would be eight times the volumn of the first octant. We first express it in polar coordinates:
    $$
        x^2 + y^2 = r^2 \implies z = \sqrt{a^2 - r^2}
    $$
    taking the partial derivatives:
    $$
        \frac{\partial z}{\partial r} = \frac{-r}{\sqrt{a^2 - r^2}}, \quad \frac{\partial z}{\partial \theta} = 0
    $$
    Thus, we can set up the double integral as follows:
    \begin{align*}
        S_{\text{1st octant}} &= \iint_R \sqrt{1 + \left( \frac{\partial z}{\partial r} \right)^2 + \left( \frac{\partial z}{\partial \theta} \right)^2} \, dA \\
        &= \int_0^{\frac{\pi}{2}} \int_0^a \sqrt{1 + \left( \frac{-r}{\sqrt{a^2 - r^2}} \right)^2 + 0^2} \, r \, dr \, d\theta \\
        &= \int_0^{\frac{\pi}{2}} \int_0^a \sqrt{1 + \frac{r^2}{a^2 - r^2}} \, r \, dr \, d\theta \\
        &= \int_0^{\frac{\pi}{2}} \int_0^a \frac{a}{\sqrt{a^2 - r^2}} \, r \, dr \, d\theta \\
        &= \int_0^{\frac{\pi}{2}} \left[ -a \sqrt{a^2 - r^2} \right]_{r=0}^{r=a} d\theta = \int_0^{\frac{\pi}{2}} a^2 \, d\theta = \frac{\pi a^2}{2}
    \end{align*}
    Therefore, the total surface area of the sphere is:
    $$        S = 8 \cdot S_{\text{1st octant}} = 4\pi a^2 $$

### 2025-09-14 YEAR2/CHE260_LectureNotes/main.tex
push

    Assuming for a Quaso-equilibrium process, and no friction, we can compute the boundary work as:
    \begin{equation}
        \delta W = F dx  = P A dx = - P dV
    \end{equation}
    where $P$ is the pressure of the gas, $A$ is the area of the piston, and $dx$ is the displacement of the piston. The negative sign is because when the gas expands, it does work on the surroundings, so the work done by the system is negative. Also, we can consider the total work by:
    \begin{equation}
        w_{12} = -\int_{V_1}^{V_2} P dV
    \end{equation}
    \text{Note} Connecting that to the $PV$ diagram, the work done by the system is the area under the curve in the $PV$ diagram.
\end{definition}

>>>>

    Assuming for a Quaso-equilibrium process, and no friction, we can compute the boundary work as:
    \begin{equation}
        \delta W = F dx  = P A dx = - P dV
    \end{equation}
    where $P$ is the pressure of the gas, $A$ is the area of the piston, and $dx$ is the displacement of the piston. The negative sign is because when the gas expands, it does work on the surroundings, so the work done by the system is negative. Also, we can consider the total work by:
    \begin{equation}
        w_{12} = -\int_{V_1}^{V_2} P dV
    \end{equation}
    \text{Note} Connecting that to the $PV$ diagram, the work done by the system is the area under the curve in the $PV$ diagram:

\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
        \draw[->] (0,0) -- (5,0) node[right] {Volume (V)};
        \draw[->] (0,0) -- (0,5) node[above] {Pressure (P)};
        \draw[thick, domain=1:4, samples=100] plot (\x, {4/\x}) node[right] {$P V = \text{constant}$};
        \fill[blue!20, opacity=0.5] (1,4) -- plot[domain=1:3] (\x, {4/\x}) -- (3, 0) -- (1, 0) -- cycle;
        \node at (2,1) {$w_{12}$};
        \node at (1,4.5) {State 1};
        \node at (3,1.5) {State 2};
    \end{tikzpicture}
    \caption{Work done by the system during expansion from State 1 to State 2 is the area under the curve in the $PV$ diagram.}
\end{figure}

\end{definition}
\newpage
\begin{example}[Isochoric Process]
    Consider a gas in a cylinder with a fixed piston. If we heat the gas, its temperature and pressure increase, but its volume remains constant. In this case, there is no boundary work done by the system, as the volume does not change ($dV = 0$). However, there may be other forms of work done on or by the system, such as electrical work if the gas is heated using an electric heater. We can demosrate by the below diagram:
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
            \draw[->] (0,0) -- (5,0) node[right] {Volume (V)};
            \draw[->] (0,0) -- (0,5) node[above] {Pressure (P)};
            \draw[thick] (2,0.5) -- (2,4) node[above] {};    
            \node at (2.25,2) {$w_{12} = 0$};
            \node at (2,4.5) {State 2};
            \node at (2,0.5) {State 1};
        \end{tikzpicture}
        \caption{Isochoric Process in a $PV$ Diagram}
    \end{figure}

\begin{example}[Isobaric Process]
    Consider a gas in a cylinder with a movable piston with mass $m$ on top. If we heat the gas, its temperature and volume increase, while its pressure remains constant (equal to the atmospheric pressure plus the pressure exerted by the mass on top of the piston). In this case, there is boundary work done by the system as the volume changes ($dV \neq 0$). The work done by the system can be calculated using the formula:
    \begin{equation}
        w_{12} = -\int_{V_1}^{V_2} P dV = -P (V_2 - V_1)
    \end{equation}
    where $P$ is the constant pressure of the gas, and $V_1$ and $V_2$ are the initial and final volumes of the gas, respectively. We can demonstrate by the below diagram:
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
            \draw[->] (0,0) -- (5,0) node[right] {Volume (V)};
            \draw[->] (0,0) -- (0,5) node[above] {Pressure (P)};
            \draw[thick] (1,2) -- (4,2) node[right] {};    
            \fill[blue!20, opacity=0.5] (1,0) -- (4,0) -- (4,2) -- (1,2) -- cycle;
            \node at (3,1) {$w_{12} = -P (V_2 - V_1)$};
            \node at (4,2.5) {State 2};
            \node at (1,1.5) {State 1};
        \end{tikzpicture}
        \caption{Isobaric Process in a $PV$ Diagram}
    \end{figure}
\end{example}

\begin{example}[Isothermal Process]
    Consider a gas in a cylinder with a movable piston. If we compress the gas very slowly while keeping it in contact with a thermal reservoir, its temperature remains constant (isothermal process). In this case, there is boundary work done by the system as the volume changes ($dV \neq 0$). The work done by the system can be calculated using the formula:
    \begin{equation}
        w_{12} = -\int_{V_1}^{V_2} P dV = -nR_u T \ln\left(\frac{V_2}{V_1}\right)
    \end{equation}
    where $n$ is the number of moles of gas, $R_u$ is the universal gas constant, $T$ is the constant temperature of the gas, and $V_1$ and $V_2$ are the initial and final volumes of the gas, respectively. We can demonstrate by the below diagram:
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
            \draw[->] (0,0) -- (5,0) node[right] {Volume (V)};
            \draw[->] (0,0) -- (0,5) node[above] {Pressure (P)};
            \draw[thick, domain=1:4, samples=100] plot (\x, {4/\x}) node[right] {$P V = \text{constant}$};
            \fill[blue!20, opacity=0.5] (1,4) -- plot[domain=1:3] (\x, {4/\x}) -- (3, 0) -- (1, 0) -- cycle;
            \node at (2,1) {$w_{12}$};
            \node at (1,4.5) {State 1};
            \node at (3,1.5) {State 2};
        \end{tikzpicture}
        \caption{Isothermal Process in a $PV$ Diagram}
    \end{figure}

    We can use the above equation to calculate the work done by the gas during the isothermal compression process.

\paragraph{Curve Fitting in the PV Diagram} In general, we can fit a curve in the $PV$ diagram using the polytropic process equation:
\begin{equation}
    P V^n = \text{constant}
\end{equation}
where $n$ is the polytropic index. Different values of $n$ correspond to different specific types of processes:
\begin{itemize}
    \item $n = 0$: Isobaric process (constant pressure)
    \item $n = 1$: Isothermal process (constant temperature)
    \item $n = \gamma$: Adiabatic process (no heat transfer, where $\gamma = \frac{c_p}{c_v}$ is the heat capacity ratio)
    \item $n \to \infty$: Isochoric process (constant volume)
\end{itemize}
We can compute the work done by the system during a polytropic process as:
\begin{equation}
    w_{12} = -\int_{V_1}^{V_2} P dV = \frac{P_2 V_2 - P_1 V_1}{n - 1} \quad (n \neq 1)
\end{equation}

\begin{definition}[Flow Work]
    Flow work is the work required to push mass into or out of a control volume. It is given by:
    \begin{equation}
        W_{flow} = \frac{PB}{m} = Pv
    \end{equation}
    where $P$ is the pressure, $B$ is the volume flow rate, and $m$ is the mass flow rate. $v$ is the specific volume (volume per unit mass).
    \textbf{Energy flowing a system per unit mass of a fluid} is called \textbf{flow energy} or \textbf{flow work}. It is the energy required to push mass into or out of a control volume. The flow work per unit mass is given by:
    \begin{equation}
       H = U + PV
    \end{equation}
    which is called the \textbf{enthalpy} of the fluid. The enthalpy is an extensive property, so we can define the specific enthalpy as:
    \begin{equation}
        h = u + Pv
    \end{equation}
\end{definition}

### 2025-09-14 YEAR2/ECE253_LectureNotes/main.tex
push

\begin{shaded}
    \paragraph{NAND and NOR Gates are Cheaper} NAND gates and NOR gates are cheaper than AND and OR gates because they require fewer transistors to implement. A 2-input NAND gate can be implemented using 4 transistors, while a 2-input AND gate requires 6 transistors (4 for the NAND gate and 2 for the NOT gate). The same applies to NOR and OR gates.

    \paragraph{NAND and NOR Gates are Universal (Functionally Complete)} Additionally, NAND and NOR gates are universal gates, meaning that any logic function can be implemented using only NAND or NOR gates.

@@

\subsection{Combinational Logic Circuits}

@@

\section{Digital Storage Elements}

>>>>

\begin{shaded}
    \textbf{NAND and NOR Gates are Cheaper} NAND gates and NOR gates are cheaper than AND and OR gates because they require fewer transistors to implement. A 2-input NAND gate can be implemented using 4 transistors, while a 2-input AND gate requires 6 transistors (4 for the NAND gate and 2 for the NOT gate). The same applies to NOR and OR gates.

    \textbf{NAND and NOR Gates are Universal (Functionally Complete)} Additionally, NAND and NOR gates are universal gates, meaning that any logic function can be implemented using only NAND or NOR gates.

@@

@@

\subsection{Boolean Algebra and Logic Minimization}
\begin{definition}[Boolean Algebra]
    Boolean algebra is a branch of algebra that deals with binary variables and logical operations. It is a effective means to describe logic circuits with a set of rules derived from the axioms of Boolean algebra.
\end{definition}

\begin{definition}[Axioms of Boolean Algebra]
    The axioms of Boolean algebra are a set of fundamental rules that govern the behavior of binary variables and logical operations. The number stems consist only of the set $\{0, 1\}$, with the following axioms:
    \begin{itemize}
        \item $0 \cdot 0 = 0$
        \item $1 \cdot 1 = 1$
        \item $0 \cdot A = 0 \cdot 1 = 1 \cdot 0 = 0$ for any $A$
        \item if $x=0$ then $\overline{x} = 1$
    \end{itemize}

    \textbf{Dual Form} We can also derive the following logical equivalences from the axioms:
    \begin{itemize}
        \item $A + 0 = A$
        \item $A + 1 = 1$
        \item $0 + 1 = 1 + 0 = 1$
        \item $A + \overline{A} = 1$
    \end{itemize}

    where $1$ is the multiplicative identity and $0$ is the additive identity.
\end{definition}

\paragraph{Rules derived from the Axioms of Boolean Algebra} The following rules can be derived from the axioms of Boolean algebra:
\begin{theorem}
\begin{itemize}
    \item $x \cdot 0 = 0$ (Annihilation)
    \item $x \cdot 1 = 1 \cdot x = x$ (Identity)
    \item $x \cdot \overline{x} = 0$ (Complementation)
    \item $x \cdot x = x$ (Idempotent)
    \item $x + 0 = 0 + x = x$ (Identity)
    \item $x + 1 = 1 + x = 1$ (Annihilation)
    \item $x + \overline{x} = 1$ (Complementation)
\end{itemize}
\end{theorem}

\begin{theorem}
The following identities can be derived from the axioms of Boolean algebra:
\begin{itemize}
    \item Commutative Laws:
    \begin{itemize}
        \item $A + B = B + A$
        \item $A \cdot B = B \cdot A$
    \end{itemize}
    \item Associative Laws:
    \begin{itemize}
        \item $A + (B + C) = (A + B) + C$
        \item $A \cdot (B \cdot C) = (A \cdot B) \cdot C$
    \end{itemize}
    \item Distributive Laws:
    \begin{itemize}
        \item $A \cdot (B + C) = A \cdot B + A \cdot C$
        \item $A + (B \cdot C) = (A + B) \cdot (A + C)$
    \end{itemize}
\end{itemize}
\end{theorem}
\begin{proof}
    By perfect induction. We can exhaustively check all possible values of $A$, $B$, and $C$ (0 or 1) to verify that both sides of each identity yield the same result.
\end{proof}

\begin{theorem}[Covering Theorem]
     The following is true:
    $$    
    x + xy = x
    $$
    and its dual:
    $$
    x(x + y) = x
    $$
\end{theorem}

\begin{theorem}[Combining Theorem]
    The following is true:
    $$
    xy + x\overline{y} = x
    $$
    and its dual:
    $$
    (x + y)(x + \overline{y}) = x
    $$

\end{theorem}

\begin{theorem}[De Morgan's Theorem]
    The following is true:
    $$
    \overline{xy} = \overline{x} + \overline{y}
    $$
    and its dual:
    $$
    \overline{x + y} = \overline{x} \cdot \overline{y}
    $$
\end{theorem}
\begin{proof}
    By direct proof. We have:
    \begin{align*}
        \overline{xy} &= \overline{x}\overline{y} + \overline{x}y + x\overline{y} \quad (\text{In Canonical SOP Form}) \\
        &= \overline{x}\overline{y} + \overline{x}y + x\overline{y} + x\overline{y} \quad (\text{Adding } x\overline{y} \text{ using $x + x = x$}) \\
        &= \overline{x}(\overline{y} + y) + \overline{y}(x + \overline{x}) \quad (\text{Using Distributive Law}) \\
        &= \overline{x} \cdot 1 + \overline{y} \cdot 1 \quad (\text{Using Complementation}) \\
        &= \overline{x} + \overline{y} \quad (\text{Using Identity})
    \end{align*}
\end{proof}

\begin{theorem}[Absorption / Redundancy Theorem]
    The following is true:
    $$
    x + \overline{x}y = x + y
    $$
    and its dual:
    $$
    x(\overline{x} + y) = xy
    $$
\end{theorem}
\begin{proof}
    By direct proof. We have:
    \begin{align*}
        x + \overline{x}y &= x + \overline{x}y + xy \quad (\text{Adding } xy \text{ using } x + xy = x) \\
        &= x(1 + y) + \overline{x}y \quad (\text{Using Distributive Law}) \\
        &= x \cdot 1 + \overline{x}y \quad (\text{Using Identity}) \\
        &= x + y \quad (\text{Using Combining Theorem})
    \end{align*}
\end{proof}

\begin{shaded}
    \textbf{Logic Minimization} The goal of logic minimization is to reduce the number of logic gates and inputs in a digital circuit while maintaining its functionality. This is important because it can lead to cost savings, improved performance, and reduced power consumption. Logic minimization can be achieved through various techniques, including Boolean algebra simplification, Karnaugh maps, and the Quine-McCluskey algorithm.
\end{shaded}

\begin{theorem}[Nand as SOP]
    And SOP circuit can be implemented using only NAND gates.
\end{theorem}

\begin{theorem}[Nor as POS]
    A POS circuit can be implemented using only NOR gates.
\end{theorem}

\begin{example}[Gumball Fact]
    Consider three sensors $s_0, s_1, s_2$ that detect defects in Gumballs. Those sensors are normally 0, but would be 1 if a defect is detected as follows:
    $$
        \begin{cases}
            s_0 = 1 & \text{if the Gumball is too small} \\
            s_1 = 1 & \text{if the Gumball is too big} \\
            s_2 = 1 & \text{if the Gumball is too light}
        \end{cases}
    $$
    We are to design a circuit that would output 1 if the Gumball is either too large or too small and too light. We can express canonical SOP form as:
    \begin{align*}
        L(s_0, s_1, s_2) &= m_3 + m_4 + m_5 + m_6 + m_7 = \overline{s_0}s_1s_2 + s_2\overline{s_0}\overline{s_2} + s_0\overline{s_1}s_2 + s_0s_1\overline{s_2} + s_0s_1s_2 \\
        \intertext{Using the Combining Theorem}
        &= \overline{s_0}s_1s_2 + s_2\overline{s_0}\overline{s_2} + s_0s_2 + s_0s_1 \\
        \intertext{Using the Absorption Theorem}
        &= s_2\overline{s_0} + s_0s_2 + s_0s_1 \\
        \intertext{Using the Covering Theorem}
        &= s_2 + s_0s_1 
    \end{align*}
\end{example}
\section{Digital Storage Elements}
